---
title: "Introduction to R and RStudio"
teaching: 45
exercises: 10
questions:
- "How to find your way around RStudio?"
- "How to interact with R?"
- "How to manage your environment?"
- "How to install packages?"
objectives:
- "Describe the purpose and use of each pane in the RStudio IDE"
- "Locate buttons and options in the RStudio IDE"
- "Define a variable"
- "Assign data to a variable"
- "Manage a workspace in an interactive R session"
- "Use mathematical and comparison operators"
- "Call functions"
- "Manage packages"
keypoints:
- "Use RStudio to write and run R programs."
- "R has the usual arithmetic operators and mathematical functions."
- "Use `<-` to assign values to variables."
- "Use `ls()` to list the variables in a program."
- "Use `rm()` to delete objects in a program."
- "Use `install.packages()` to install packages (libraries)."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options-short.R")
knitr_fig_path("01-")
```


## Introduction to RStudio
fundamentals of R, best practices that will make your life easier

explain R studio panels (editor appears later)
all coding goes in script

you type in commands, R tries to execute them, and then returns a result.

# Using R as a calculator

The simplest thing you could do with R is do arithmetic:

```{r}
1 + 100
```

explain results `[1]`

explain leading `>` vs `+`

aborting: `Esc` ; `ctrl+c`

```{r}
# comments are not executed
```

order of operations

```{r}
3 + 5 * 2
```

parentheses

```{r}
(3 + 5) * 2
```

Really small or large numbers -> scientific notation

```{r}
2^300
(1+1)^300
2/10000
```

Which is shorthand for "multiplied by `10^XX`". So `2e-4` is shorthand for `2 * 10^(-4)`.

You can write numbers in scientific notation too:

```{r}
5e3  # Note the lack of minus here
```

## Comparing things

We can also do comparison in R:

```{r}
1 == 1  
1 != 2  
```

```{r}
1 <  2  
1 <= 1 
1 > 0
1 >= -9 
```

# CONCEPT: Variables

let's say we want to take some important statistics on the class today

raise your hand if you prefer cats over dogs ... dogs over cats ... 

```{r}
# don't write comments
12 # cats over dogs
16 # dogs over cats
32 # total
```

if we come back tomorrow, will we remember what this means? no way!
so we want to store these values in a way that makes it easy to remember what they represent and easy to reuse them

## Variables and assignment

```{r}
preferCats <- 12
```

does not print a value

See `Environment` tab 

print value by 
```{r}
preferCats
```

store others

```{r}

preferDogs <- 16 
totalInClass <- 31
noPreference <- totalInClass - preferCats - preferDogs

diffInPreferences <- preferCats - preferDogs # NOTE: SMALLER - BIGGER

```


variables can be reassigned

```{r}

totalInClass <- 32

preferCats <- preferCats + 1
```

see updated variable pane

```{r}
# note that this is not updated!
diffInPreferences
```

**explain order of commands**

# CONCEPT: Functions

```{r}
diffInPreferences
```

this is a function, takes a parameter

```{r}

diffInPreferences
# I don't want a negative
abs(diffInPreferences)

abs(-10)

abs

?abs

```

hard to remember names -- use google
code completion

# Misc

## flexibility of language


*Variable names* can contain letters, numbers, underscores and periods. They
cannot start with a number nor contain spaces. 

  * camelCaseToSeparateWords
  * periods.between.words
  * underscores\_between_words

*value assignment symbol*

```{r}
totalInClass = 32
```

recommendation is to use `<-`



What you use is up to you, but **be consistent**.


## R Packages

package is a group of functions
over 7,000 packages available on CRAN (the comprehensive R archive
network).

* You can see what packages are installed by typing
  `installed.packages()`
* You can install packages by typing `install.packages("packagename")`,
  where `packagename` is the package name, in quotes.
* You can make a package available for use with `library(packagename)`


> ## Challenge 1 {.challenge}
>
> What will be the value of each  variable  after each
> statement in the following program?
>
> ```{r, eval=FALSE}
> mass <- 47.5
> age <- 122
> mass <- mass * 2.3
> age <- age + 1
> ```
>

> ## Challenge 2 {.challenge}
>
> Run the code from the previous challenge, and write a command to
> compare mass to age. Is mass larger than age?
>

> ## Challenge 3 {.challenge}
>
> Clean up your working environment by deleting the mass and age
> variables.
>

> ## Challenge 4 {.challenge}
> 
> Install the following packages: `ggplot2`, `plyr`, `gapminder`
> 
