---
title: "Project Management With RStudio"
teaching: 20
exercises: 10
questions:
- "How can I manage my projects in R?"
objectives:
- Create self-contained projects in RStudio
keypoints:
- "Use RStudio to create and manage projects with consistent layout."
- "Treat raw data as read-only."
- "Treat generated output as disposable."
- "Separate function definition and application."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options-short.R")
knitr_fig_path("02-")
```


## Introduction

The scientific process is naturally incremental

we want to avoid a folder filled with 200 files

# Make a project

```{r}
# make new project
```

> ## Tip: Good Enough Practices for Scientific Computing
>
> [Good Enough Practices for Scientific Computing](https://github.com/swcarpentry/good-enough-practices-in-scientific-computing/blob/gh-pages/good-enough-practices-for-scientific-computing.pdf) gives the following recommendations for project organization:
>
> 1. Put each project in its own directory, which is named after the project.
> 2. Put text documents associated with the project in the `doc` directory.
> 3. Put raw data and metadata in the `data` directory, and files generated during cleanup and analysis in a `results` directory.
> 4. Put source for the project's scripts and programs in the `src` directory, and programs brought in from elsewhere or compiled locally in the `bin` directory.
> 5. Name all files to reflect their content or function.
>
{: .callout}

## Best practices for project organisation

## Raw data = read only

Always keep the original version of your data - never edit this directly

## Data Cleaning/Munging

use a script, store the results

## Treat generated output as disposable

output folder with relevant subfolders

R folder with functions you create

# Create directories

data
R
outputs


## Save the data in the data directory

Now we have a good directory structure we will now place/save the data file in the `data/` directory.

```{r}

# save this script

# function with named parameters
#"https://raw.githubusercontent.com/theavanrossum/r-novice-gapminder/gh-pages/data/feline-data.csv"
download.file(url = "https://bit.ly/2pD8vEF", destfile = "../data/feline-data.csv")

```

> ## Challenge 1 {.challenge}
> Download the gapminder data from [here](https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv).
>
> 1. Download the file (CTRL + S, right mouse click -> "Save as", or File -> "Save page as")
> 2. Make sure it's saved under the name `gapminder-FiveYearData.csv`
> 3. Save the file in the `data/` folder within your project.
>
> We will load and inspect these data later.


