---
title: "Data Structures"
teaching: 40
exercises: 15
questions:
- "How can I read data in R?"
- "What are the basic data types in R?"
- "How do I represent categorical information in R?"
objectives:
- "To be aware of the different types of data."
- "To begin exploring data frames, and understand how they are related to vectors, factors and lists."
- "To be able to ask questions from R about the type, class, and structure of an object."
keypoints:
- "Use `read.csv` to read tabular data in R."
- "The basic data types in R are double, integer, complex, logical, and character."
- "Use factors to represent categories in R."
source: Rmd
---

We saw how to use numbers by storing them as named variables, but what if we want to store whole datasets or words or a list of numbers? then we need to look at R's data types and data structures

*data type* -- type of data

*data structure* -- structure that holds data

```{r, include=FALSE}
source("../bin/chunk-options-short.R")
knitr_fig_path("04-")

```


1st data structure: data frames

# Reading in data

```{r}
cats <- read.csv(file = "../data/feline-data.csv")
# read.delim
# read.table
```

```{r}
cats # explain DFs
```

We can see that it is a *data.frame* by calling the `class` function on it:

```{r}
class(cats)
```

```{r}
cats$weight
cats$coat
```

We can do other operations on the columns:

```{r}
## Say we discovered that the scale weighs two Kg light:
cats$weight + 2
cats$weight + cats$likes_string
```

But what about

```{r}
cats$weight + cats$coat
```

Understanding what happened here is key to successfully analyzing data in R.

## Data Types

 `2.1` + `"black"` is nonsense -- *data types* can be intuitive

```{r}
typeof(cats$weight)
```

There are 5 main types: `double`, `integer`, `complex`, `logical` and `character`.

```{r}
typeof(3.14)
typeof(1L) # The L suffix forces the number to be an integer, since by default R uses float numbers
typeof(1+1i)
typeof(TRUE)
typeof('banana')
```

```{r}
typeof(cats$height)
```

```{r}
cats$height + 2
```

What happened?

everything in a column be the same basic type

## Vectors and Type Coercion

brings us to our second data structure: the *vector*.
A vector: ordered list of things, *everything is the same basic data type*

```{r}

myVector <- vector(mode='character', length=3)

myVector
```

You can check if something is a vector:

```{r}
typeof(myVector) # data type
str(myVector) # data structure

```

```{r}
str(cats$weight)
```

data frames are a list of vectors - each column in a dataframe is a vector

allows us to make assumptions about content of column for operations

```{r}
combine_vector <- c(2,6,3) # combine function
combine_vector
```

> ## Challenge (can skip)
make two vectors, each with 4 integers so that when you sum them, the values in the resultant vector sums to 10


Given what we've learned so far, what do you think the following will produce?

```{r}
quiz_vector <- c(2,6,'3')
```

*type coercion*: mix of types in vector -> force them all to be the same type

```{r}
coercion_vector <- c('a', TRUE)
coercion_vector
another_coercion_vector <- c(0, TRUE)
another_coercion_vector
```

`logical` -> `integer` -> `numeric` -> `complex` -> `character`

 -> = *are transformed into*. 

force coercion 

```{r}
character_vector_example <- c('0','2','4')
character_vector_example
character_coerced_to_numeric <- as.numeric(character_vector_example)
character_coerced_to_numeric
numeric_coerced_to_logical <- as.logical(character_coerced_to_numeric)
numeric_coerced_to_logical
```

if you get unexpected results - check your data types

But coercion can also be very useful! 
We should use the `logical` datatype here, which has two states: `TRUE` or `FALSE`

```{r}
cats$likes_string
cats$likes_string <- as.logical(cats$likes_string)
cats$likes_string
```

The combine function, `c()`, will also append things to an existing vector:

```{r}
ab_vector <- c('a', 'b')
ab_vector
combine_example <- c(ab_vector, 'SWC')
combine_example
```

You can also make series of numbers:

```{r}
mySeries <- 1:10
mySeries
```

We can ask a few questions about vectors:

```{r}
sequence_example <- 1:10
head(x = sequence_example, n=2)
tail(x = sequence_example, n=4)
length(sequence_example)
class(sequence_example)
typeof(sequence_example)
str(sequence_example)
```

Finally, you can give names to elements in your vector:

```{r}
my_example <- 5:8
names(my_example) <- c("a", "b", "c", "d")
my_example
names(my_example)
```

> ## Challenge 1.1
>
> Start by making a vector with the numbers 1 through 26.
> Multiply the vector by 2, and give the resulting vector
> names A through Z (hint: there is a built in vector called `LETTERS`)
>
> > ## Solution to Challenge 1.1
> >
> > ```{r}
> > x <- 1:26
> > x <- x * 2
> > names(x) <- LETTERS
> > ```
> {: .solution}
{: .challenge}


## Factors

We said that columns in data.frames were vectors:

```{r}
str(cats$weight)
str(cats$likes_string)
```

These make sense. But what about

```{r}
str(cats$coat)
```

*factor*: categorical information

```{r}
coats <- c('tabby', 'tortoiseshell', 'tortoiseshell', 'black', 'tabby')
coats
str(coats)
```

We can turn a vector into a factor like so:

```{r}
CATegories <- as.factor(coats)
class(CATegories)
CATegories
str(CATegories)
```

stored as numbers

```{r}
typeof(CATegories)
```

> ## Challenge 2
>
> Is there a factor in our `cats` data.frame? what is its name?
> Try using `?read.csv` to figure out how to keep text columns as character
> vectors instead of factors; then write a command or two to show that the factor
> in `cats` is actually a character vector when loaded in this way.
>
> > ## Solution to Challenge 2
> >
> > One solution is use the argument `stringAsFactors`:
> >
> > ```{r, eval = FALSE}
> > cats <- read.csv(file="data/feline-data.csv", stringsAsFactors=FALSE)
> > str(cats$coat)
> > ```
> >
> > Another solution is use the argument `colClasses`
> > that allow finer control.
> >
> > ```{r, eval = FALSE}
> > cats <- read.csv(file="data/feline-data.csv", colClasses=c(NA, NA, "character"))
> > str(cats$coat)
> > ```
> >
> > Note: new students find the help files difficult to understand; make sure to let them know
> > that this is typical, and encourage them to take their best guess based on semantic meaning,
> > even if they aren't sure.
> {: .solution}
{: .challenge}

In modelling functions, it's important to know what the baseline levels are. This
is assumed to be the first factor, but by default factors are labelled in
alphabetical order. You can change this by specifying the levels:

```{r}
mydata <- c("case", "control", "control", "case")
factor_ordering_example <- factor(mydata, levels = c("control", "case"), ordered=TRUE)
str(factor_ordering_example)
```

In this case, we've explicitly told R that "control" should represented by 1, and
"case" by 2. This designation can be very important for interpreting the
results of statistical models!

## Lists

Another data structure you'll want in your bag of tricks is the `list`. A list
is simpler in some ways than the other types, because you can put anything you
want in it:

```{r}
list_example <- list(1, "a", TRUE, 1+4i)
list_example
another_list <- list(title = "Numbers", numbers = 1:10, data = TRUE )
another_list
```

We can now understand something a bit surprising in our data.frame; what happens if we run:

```{r}
typeof(cats)
```

data.frame is really a list of vectors and factors

therefore 

*columns must have the same type*
*rows do not have to*

> ## Challenge 3
>
> There are several subtly different ways to call variables, observations and
> elements from data.frames:
>
> - `cats[1]`
> - `cats[[1]]`
> - `cats$coat`
> - `cats["coat"]`
> - `cats[1, 1]`
> - `cats[, 1]`
> - `cats[1, ]`
>
> Try out these examples and explain what is returned by each one.
>
> *Hint:* Use the function `typeof()` to examine what is returned in each case.
>
> > ## Solution to Challenge 3
> > ```{r, eval=TRUE, echo=TRUE}
> > cats <- read.csv(file="../data/feline-data.csv", stringsAsFactors=FALSE)
> > cats[1]
> > ```
> > We can think of a data frame as a list of vectors. The single brace `[1]`
> returns the first slice of the list, as another list. In this case it is the
> first column of the data frame.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[[1]]
> > ```
> > The double brace `[[1]]` returns the contents of the list item. In this case
> it is the contents of the first column, a _vector_ of type _factor_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats$coat
> > ```
> > This example uses the `$` character to address items by name. _coat_ is the
> first column of the data frame, again a _vector_ of type _factor_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats["coat"]
> > ```
> > Here we are using a single brace `["coat"]` replacing the index number with
> the column name. Like example 1, the returned object is a _list_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[1, 1]
> > ```
> > This example uses a single brace, but this time we provide row and column
> coordinates. The returned object is the value in row 1, column 1. The object
> is an _integer_ but because it is part of a _vector_ of type _factor_, R
> displays the label "calico" associated with the integer value.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[, 1]
> > ```
> > Like the previous example we use single braces and provide row and column
> coordinates. The row coordinate is not specified, R interprets this missing
> value as all the elements in this _column_ _vector_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[1, ]
> > ```
> > Again we use the single brace with row and column coordinates. The column
> coordinate is not specified. The return value is a _list_ containing all the
> values in the first row.
> {: .solution}
{: .challenge}

## Matrices

Last but not least is the matrix. We can declare a matrix full of zeros:

```{r}
matrix_example <- matrix(data = c(0), ncol=6, nrow=3)
matrix_example
```

And similar to other data structures, we can ask things about our matrix:

```{r}
class(matrix_example)
str(matrix_example)

dim(matrix_example)
nrow(matrix_example)
ncol(matrix_example)
```

> ## Challenge 4
>
> What do you think will be the result of
> `length(matrix_example)`?
> Try it.
> Were you right? Why / why not?
>
> > ## Solution to Challenge 4
> >
> > What do you think will be the result of
> > `length(matrix_example)`?
> >
> > ```{r}
> > matrix_example <- matrix(0, ncol=6, nrow=3)
> > length(matrix_example)
> > ```
> >
> > Because a matrix is a vector with added dimension attributes, `length`
> > gives you the total number of elements in the matrix.
> {: .solution}
{: .challenge}


> ## Challenge 5
>
> Make another matrix, this time containing the numbers 1:50,
> with 5 columns and 10 rows.
> Did the `matrix` function fill your matrix by column, or by
> row, as its default behaviour?
> See if you can figure out how to change this.
> (hint: read the documentation for `matrix`!)
>
> > ## Solution to Challenge 5
> >
> > Make another matrix, this time containing the numbers 1:50,
> > with 5 columns and 10 rows.
> > Did the `matrix` function fill your matrix by column, or by
> > row, as its default behaviour?
> > See if you can figure out how to change this.
> > (hint: read the documentation for `matrix`!)
> >
> > ```{r, eval=FALSE}
> > x <- matrix(1:50, ncol=5, nrow=10)
> > x <- matrix(1:50, ncol=5, nrow=10, byrow = TRUE) # to fill by row
> > ```
> {: .solution}
{: .challenge}


> ## Challenge 6
>  Create a list of length two containing a character vector for each of the sections in this part of the workshop:
>
>  - Data types
>  - Data structures
>
>  Populate each character vector with the names of the data types and data
>  structures we've seen so far.
>
> > ## Solution to Challenge 6
> > ```{r}
> > dataTypes <- c('double', 'complex', 'integer', 'character', 'logical')
> > dataStructures <- c('data.frame', 'vector', 'factor', 'list', 'matrix')
> > answer <- list(dataTypes, dataStructures)
> > ```
> > Note: it's nice to make a list in big writing on the board or taped to the wall
> > listing all of these types and structures - leave it up for the rest of the workshop
> > to remind people of the importance of these basics.
> >
> {: .solution}
{: .challenge}


> ## Challenge 7
>
> Consider the R output of the matrix below:
> ```{r, echo=FALSE}
> matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
> ```
> What was the correct command used to write this matrix? Examine
> each command and try to figure out the correct one before typing them.
> Think about what matrices the other commands will produce.
>
> 1. `matrix(c(4, 1, 9, 5, 10, 7), nrow = 3)`
> 2. `matrix(c(4, 9, 10, 1, 5, 7), ncol = 2, byrow = TRUE)`
> 3. `matrix(c(4, 9, 10, 1, 5, 7), nrow = 2)`
> 4. `matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)`
>
> > ## Solution to Challenge 7
> >
> > Consider the R output of the matrix below:
> > ```{r, echo=FALSE}
> > matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
> > ```
> > What was the correct command used to write this matrix? Examine
> > each command and try to figure out the correct one before typing them.
> > Think about what matrices the other commands will produce.
> > ```{r, eval=FALSE}
> > matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
> > ```
> {: .solution}
{: .challenge}
