---
title: Vectorization
teaching: 10
exercises: 15
questions:
- "How can I operate on all the elements of a vector at once?"
objectives:
- "To understand vectorized operations in R."
keypoints:
- "Use vectorized operations instead of loops."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options-short.R")
knitr_fig_path("09-")

```

Most of R's functions are vectorized: function will
operate on all elements of a vector

```{r}
x <- 1:4
x * 2

x > 2 # Comparison operators

log(x) # functions
```

```{r}
y <- 6:9
x + y
```

```{r}
m <- matrix(1:12, nrow=3, ncol=4)
m * -1

```


*matrix algebra* has its own symbols  `%*%`

> ## Challenge 3
>
> Given the following matrix:
>
> ```{r}
> m <- matrix(1:12, nrow=3, ncol=4)
> m
> ```
>
> Write down what you think will happen when you run:
>
> 1. `m ^ -1`
> 2. `m * c(1, 0, -1)`
> 3. `m > c(0, 20)`
> 4. `m * c(1, 0, -1, 2)`
>
> Did you get the output you expected? If not, ask a helper!
>
> > ## Solution to challenge 3
> >
> > Given the following matrix:
> >
> > ```{r}
> > m <- matrix(1:12, nrow=3, ncol=4)
> > m
> > ```
> >
> >
> > Write down what you think will happen when you run:
> >
> > 1. `m ^ -1`
> >
> > ```{r, echo=FALSE}
> > m ^ -1
> > ```
> >
> > 2. `m * c(1, 0, -1)`
> >
> > ```{r, echo=FALSE}
> > m * c(1, 0, -1)
> > ```
> >
> > 3. `m > c(0, 20)`
> >
> > ```{r, echo=FALSE}
> > m > c(0, 20)
> > ```
> >
> {: .solution}
{: .challenge}


> ## Challenge 4
>
> We're interested in looking at the sum of the
> following sequence of fractions:
>
> ```{r, eval=FALSE}
>  x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
> ```
>
> This would be tedious to type out, and impossible for high values of
> n.  Use vectorisation to compute x when n=100. What is the sum when
> n=10,000?
>
> > ##  Challenge 4
> >
> > We're interested in looking at the sum of the
> > following sequence of fractions:
> >
> > ```{r, eval=FALSE}
> >  x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
> > ```
> >
> > This would be tedious to type out, and impossible for
> > high values of n.
> > Can you use vectorisation to compute x, when n=100?
> > How about when n=10,000?
> >
> > ```{r}
> > sum(1/(1:100)^2)
> > sum(1/(1:1e04)^2)
> > n <- 10000
> > sum(1/(1:n)^2)
> > ```
> >
> > We can also obtain the same results using a function:
> > ```{r}
> > inverse_sum_of_squares <- function(n) {
> >   sum(1/(1:n)^2)
> > }
> > inverse_sum_of_squares(100)
> > inverse_sum_of_squares(10000)
> > n <- 10000
> > inverse_sum_of_squares(n)
> > ```
> >
> {: .solution}
{: .challenge}
