---
title: "Exploring Data Frames"
teaching: 20
exercises: 10
questions:
- "How can I manipulate a data frame?"
objectives:
- "Add and remove rows or columns."
- "Remove rows with `NA` values."
- "Append two data frames."
- "Understand what a `factor` is."
- "Convert a `factor` to a `character` vector and vice versa."
- "Display basic properties of data frames including size and class of the columns, names, and first few rows."
keypoints:
- "Use `cbind()` to add a new column to a data frame."
- "Use `rbind()` to add a new row to a data frame."
- "Remove rows from a data frame."
- "Use `na.omit()` to remove rows from a data frame with `NA` values."
- "Use `levels()` and `as.character()` to explore and manipulate factors."
- "Use `str()`, `nrow()`, `ncol()`, `dim()`, `colnames()`, `rownames()`, `head()`, and `typeof()` to understand the structure of a data frame."
- "Read in a csv file using `read.csv()`."
- "Understand what `length()` of a data frame represents."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options-short.R")
knitr_fig_path("05-")
```

At this point, you've seen it all: in the last lesson, we toured all the basic
data types and data structures in R. Everything you do will be a manipulation of
those tools. But most of the time, the star of the show is the data frame—the table that we created by loading information from a csv file. In this lesson, we'll learn a few more things
about working with data frames.

## Adding columns and rows in data frames

```{r, echo = FALSE}
cats <- read.csv("../data/feline-data.csv")
```

adding a new column
columns are vectors

```{r}
age <- c(2, 3, 5, 2)
cats

cbind(cats, age)
cats
```


```{r, error=TRUE}
# too many, will fail
age <- c(2, 3, 5, 12, 23, 45) 
cbind(cats, age)
nrow(cats)
length(age)

# note vector re-use as we saw before
age <- c(2, 3)
cbind(cats, age) 

```


```{r}
age <- c(2, 3, 5, 2)
cats <- cbind(cats, age)
```

Now how about adding rows? We already know that the rows of a
data frame are lists:

```{r}
newRow <- list("R11","tortoiseshell", 3.3, TRUE, 29, 9)
cats <- rbind(cats, newRow)
cats
```

## Factors

```{r}
levels(cats$coat)
levels(cats$coat) <- c(levels(cats$coat), "tortoiseshell")
cats <- rbind(cats, list("R11","tortoiseshell", 3.3, TRUE, 29, 9))
```

Alternatively, we can change a factor into a character vector; we lose the
handy categories of the factor, but we can subsequently add any word we want to the
column without babysitting the factor levels:

```{r}
str(cats)
cats$coat <- as.character(cats$coat)
str(cats)
```

> ## Challenge 1
> Let's imagine that 1 human year is equivalent to 7 cat years. 
> 1. Create a vector called `human_age` by multiplying `cats$age` by 7.
> 2. Convert `human_age` to a factor.
> 3. Convert `human_age` back to a numeric vector using the `as.numeric()` function. Now divide it by 7 to get the original ages back. Explain what happened.
>
> > ## Solution to Challenge 1
> > 1. `human_age <- cats$age * 7`
> > 2. `human_age <- factor(human_age)`. `as.factor(human_age)` works just as well.
> > 3. `as.numeric(human_age)` yields `1 2 3 4 4` because factors are stored as integers (here, 1:4), each of which is associated with a label (here, 28, 35, 56, and 63). Converting the factor to a numeric vector gives us the underlying integers, not the labels. If we want the original numbers, we need to convert `human_age` to a character vector and then to a numeric vector (why does this work?). This comes up in real life when we accidentally include a character somewhere in a column of a .csv file supposed to only contain numbers, and forget to set `stringsAsFactors=FALSE` when we read in the data.
> {: .solution}
{: .challenge}

## Removing rows

We now know how to add rows and columns to our data frame in R—but in our
first attempt to add a "tortoiseshell" cat to the data frame we have accidentally
added a garbage row:

```{r}
cats
```

We can ask for a data frame minus this offending row:

```{r}
cats[-4, ]
```

Notice the comma with nothing after it to indicate that we want to drop the entire fourth row.

Note: we could also remove both new rows at once by putting the row numbers
inside of a vector: `cats[c(-4,-5), ]`

Alternatively, we can drop all rows with `NA` values:

```{r}
na.omit(cats)
```

Let's reassign the output to `cats`, so that our changes will be permanent:

```{r}
cats <- na.omit(cats)
```

## Removing columns  

We can also remove columns in our data frame. What if we want to remove the column "age". We can remove it in two ways, by variable number or by index.  

```{r}
cats[,-4]
```

Notice the comma with nothing before it, indicating we want to keep all of the rows.  

## Appending to a data frame

The key to remember when adding data to a data frame is that *columns are
vectors and rows are lists.* We can also glue two data frames
together with `rbind`:

```{r}
moreCats <- rbind(cats, cats)
moreCats
```


> ## Challenge 2
>
> You can create a new data frame right from within R with the following syntax:
> ```{r}
> df <- data.frame(id = c("a", "b", "c"),
>                  x = 1:3,
>                  y = c(TRUE, TRUE, FALSE),
>                  stringsAsFactors = FALSE)
> ```
> Make a data frame that holds the following information for yourself:
>
> - first name
> - last name
> - lucky number
>
> Then use `rbind` to add an entry for the people sitting beside you.
> Finally, use `cbind` to add a column with each person's answer to the question, "Is it time for coffee break?"
>
> > ## Solution to Challenge 2
> > ```{r}
> > df <- data.frame(first = c("Grace"),
> >                  last = c("Hopper"),
> >                  lucky_number = c(0),
> >                  stringsAsFactors = FALSE)
> > df <- rbind(df, list("Marie", "Curie", 238) )
> > df <- cbind(df, coffeetime = c(TRUE,TRUE))
> > ```
> {: .solution}
{: .challenge}

## Realistic example
So far, you have seen the basics of manipulating data frames with our cat data;
now let's use those skills to digest a more realistic dataset. Let's read in the
`gapminder` dataset that we downloaded previously:

```{r}
#"https://raw.githubusercontent.com/theavanrossum/r-novice-gapminder/gh-pages/data/gapminder_data.csv"
download.file(url = "https://bit.ly/2xIXR2B" ,
              destfile = "../data/gapminder_data.csv")

gapminder <- read.csv("../data/gapminder_data.csv")
```

readxl (https://cran.r-project.org/web/packages/readxl/index.html) package.


investigate gapminder

```{r}
str(gapminder)

typeof(gapminder$year)
typeof(gapminder$country)
str(gapminder$country)
```

what do you think the following will produce, and why?

```{r}
length(gapminder)
```

*df is a list*

```{r}
typeof(gapminder)
```

```{r}
nrow(gapminder)
ncol(gapminder)
dim(gapminder)
```

```{r}
colnames(gapminder)
```

data types ok?

```{r}
head(gapminder)
```

> ## Challenge 3
>
> It's good practice to also check the last few lines of your data and some in the middle. How would you do this?
>
> Searching for ones specifically in the middle isn't too hard or we could simply ask for a few lines at random. How would you code this? (hint: check out the sample() or runif() functions using ?sample )
>
> > ## Solution to Challenge 3
> > To check the last few lines it's relatively simple as R already has a function for this:
> > 
> > ~~~
> > tail(gapminder)
> > tail(gapminder, n = 15)
> > ~~~
> > {: .r}
> > 
> > What about a few arbitrary rows just for sanity (or insanity depending on your view)?
> > ## Tip: There are several ways to achieve this.
> > The solution here presents one form using nested functions. i.e. a function passed as an argument to another function. This might sound like a new concept but you are already using it in fact.
> > Remember my_dataframe[rows, cols] will print to screen your data frame with the number of rows and columns you asked for (although you might have asked for a range or named columns for example). How would you get the last row if you don't know how many rows your data frame has? R has a function for this. What about getting a (pseudorandom) sample? R also has a function for this.
> > ~~~
> > gapminder[sample(nrow(gapminder), 5), ]
> > ~~~
> > {: .r}
> {: .solution}
{: .challenge}


> ## Challenge 5
>
> Read the output of `str(gapminder)` again;
> this time, use what you've learned about factors, lists and vectors,
> as well as the output of functions like `colnames` and `dim`
> to explain what everything that `str` prints out for gapminder means.
> If there are any parts you can't interpret, discuss with your neighbors!
>
> > ## Solution to Challenge 5
> >
> > The object `gapminder` is a data frame with columns
> > - `country` and `continent` are factors.
> > - `year` is an integer vector.
> > - `pop`, `lifeExp`, and `gdpPercap` are numeric vectors.
> >
> {: .solution}
{: .challenge}
